---
# Title: Cyclistics Data Analysis
# Author: Kareem Mohamed
# Date : 15-3-2024
---

# Lodaing Packages

```{r}
library(tidyverse)
library(readr)
library(lubridate)
library(skimr)
library(janitor)
```

# Import Data 

Importing the data that i downloaded and combining it in one data frame called combined_bike_data, the data frame will combine 12 csv files containing historical data about the past 12 month of Chicago city bike riders starting from February 2024 going back to March 2023.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/Coursera Data Analysis/Cyclistic")
```

As we can see after combining the data the total observations are {5,707,168} and it has {15} variables.

```{r} 
combined_bike_data <- rbind(
read.csv("february24-divvy-tripdata.csv"),
read.csv("january24-divvy-tripdata.csv"),
read.csv("december23-divvy-tripdata.csv"),
read.csv("november23-divvy-tripdata.csv"),
read.csv("october23-divvy-tripdata.csv"),
read.csv("september23-divvy-tripdata.csv"),
read.csv("august23-divvy-tripdata.csv"),
read.csv("july23-divvy-tripdata.csv"),
read.csv("june23-divvy-tripdata.csv"),
read.csv("may23-divvy-tripdata.csv"),
read.csv("april23-divvy-tripdata.csv"),
read.csv("march23-divvy-tripdata.csv")
)
```

# Process

### Inspecting

```{r}
glimpse(combined_bike_data)
colnames(combined_bike_data)
```

### Adding N/A to empty cells

Some cells in the data frame are empty they don't have N/A values so in order to delete them or to acknowledge them i will add `NA` to them.

```{r}
combined_bike_data$start_station_name[combined_bike_data$start_station_name == ""] <- NA
combined_bike_data$start_station_id[combined_bike_data$start_station_id == ""] <- NA
combined_bike_data$end_station_name[combined_bike_data$end_station_name == ""] <- NA
combined_bike_data$end_station_id[combined_bike_data$end_station_id == ""] <- NA
```

### Adding columns

#### Date, Month, Day, and year
  Adding column for date, month, day, and year by using as.`Date()` on started_at to get the `date` and then getting the rest of the new columns.

```{r}
combined_bike_data$date <- as.Date(combined_bike_data$started_at)
combined_bike_data$month <-  format(as.Date(combined_bike_data$date), "%m")
combined_bike_data$day <- format(as.Date(combined_bike_data$date), "%d")
combined_bike_data$year <- format(as.Date(combined_bike_data$date), "%Y")
```

#### Ride Length

First im going to identify the correct format of `started_at` and `ended_at`

```{r}
unique(combined_bike_data$started_at)
```

As you can see that the format is YYY-MM-DD HH:MM:SS.
changing the data type in `started_at` and `ended_at` to the right format.


```{r}
combined_bike_data$started_at <- as.POSIXct(combined_bike_data$started_at, format = "%Y-%m-%d %H:%M:%S")

combined_bike_data$ended_at <- as.POSIXct(combined_bike_data$ended_at, format = "%Y-%m-%d %H:%M:%S")
``` 

Adding a column for ride length by deducting ended_at from started_at

```{r}
combined_bike_data <- combined_bike_data %>% mutate(ride_length = ended_at - started_at)
```

dividing the ride_length by 60 to represent it by minutes not seconds.

```{r}
combined_bike_data$ride_length <- as.numeric(combined_bike_data$ride_length)
combined_bike_data$ride_length <- as.numeric(combined_bike_data$ride_length/60)
```

#### Days of The Week
Adding column day_of_week to know the days of the week that the rides started_at.

```{r}
combined_bike_data <- combined_bike_data %>%
  mutate(day_of_week = weekdays(as.Date(combined_bike_data$started_at)))
```

### Removing Columns 

Removing data that will not be useful to the analysis so i removed `start_lat`, `start_lng`, `end_lat` and `end_lng`.

```{r}
combined_bike_data <- combined_bike_data %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))
```


### Adding N/A to specific cells  

Some cells need to be removed so i will add NA to them based on some specific conditions.
First `ride_length` it has some values that are equal to `0` and negative number that cannot be displayed  which means that the rider hasn't used the bike for any hours ,minutes or seconds and it is wrong to have the ride length in negative as well so i will not be using these cells in my analysis.

Then using `sum(is.na)` to know how many ride_length where equal to 0 and negative numbers, as you can see we had {1560} entry.

```{r}
combined_bike_data$ride_length[combined_bike_data$ride_length <= "0"] <- NA
sum(is.na(combined_bike_data$ride_length))
```
Then i will subset the data to do further adjustments to the data 

```{r}
subset_data <- combined_bike_data[!is.na(combined_bike_data$ride_length), ]
```


Next adding NA to values that are more than one day because we have some values in ride_length that are more than 12 thousand minutes which is not normal so any values that is more than one day will be replaced by NA.
 ride_length that is more than 1440 minutes = one day.

Using sum(is.na) i found that the values that are more than 1440 minutes are 6882 entry.

```{r}
subset_data[subset_data$ride_length > 1440, "ride_length"] <- NA

sum(is.na(subset_data$ride_length))
```
combining the subsets back to the original data.
and using sum(is.na) the total N/A's in the ride_length of the combined_bike_data is 8442 entry.

```{r}
combined_bike_data[!is.na(combined_bike_data$ride_length), ] <- subset_data
sum(is.na(combined_bike_data$ride_length))
```
Dropping the subset data.

```{r}
rm(subset_data)
```

### Deleting N/A Values

Using `na.omit()` to delete the `N/A` values in the data frame.

```{r}
combined_bike_data <- na.omit(combined_bike_data)
```

There were {1,375,436} observations deleted and now the available observations that i will use for analysis is {4,331,652} observation.
mostly the deleted N/A's were from the `start_station_name`, `end_station_name`, `start_station_id`, `end_station_id` and some rows in the `ride_length`.
 

# Analyze

### Summary 

using mean and min and max on the data frame to get insights.
mean is {15.9} minutes , min is {0.016} and max is {1439.8} minutes.

```{r}
mean(combined_bike_data$ride_length) 
min(combined_bike_data$ride_length)
max(combined_bike_data$ride_length)
```

Doing a summary to see what type of ride is being used more when the ride length is more than one hour and less than one hour.

First condition is using the electric and classic bikes when the length is more than 60 minutes. 
the outcome is that {298,193} rides were electric bikes and {494,360} rides were classic bikes.

Second condition is using the same rideable types when the ride length is shorter than 60 minutes.
the outcome is that {1,234,413} rides were by electric bikes and {2,232,262} were by classic bikes.
Conclusion is that the classic bikes are being used more when the ride length is less or more than 60 minutes.

```{r}
sum(combined_bike_data$rideable_type == "electric_bike" & combined_bike_data$ride_length > "60")
sum(combined_bike_data$rideable_type == "classic_bike" & combined_bike_data$ride_length > "60")
sum(combined_bike_data$rideable_type == "electric_bike" & combined_bike_data$ride_length < "60")
sum(combined_bike_data$rideable_type == "classic_bike" & combined_bike_data$ride_length < "60")
```

Using sum to know how many members do we have and how many casual riders do we have throughout the past 12 month.
we have {2,806,688} members and {1,524,848} are casual riders.

```{r}
sum(combined_bike_data$member_casual == "member")
sum(combined_bike_data$member_casual == "casual")
```

By using sum again you can see that the total electric bike used from the total rides {1,532,620} were using electric bikes and {2,726,663} were using classic bike so the classic bikes are being used more .
```{r}
sum(combined_bike_data$rideable_type == "electric_bike")
sum(combined_bike_data$rideable_type == "classic_bike")
```
using sum to see who uses electric bike and classic bike more, is it members or is it casual riders?

First we can see that the member use more electric bike than the casual riders as {961, 489} rides were by members and {571,131} rides were by casual riders.
Second we can see that members also use more classic bikes than casual riders as {1,845,199} rides were by members and {881,464} rides were by casual riders.

We can see as well that casual riders and members tend to use classic bikes more than electric bikes.

```{r}
sum(combined_bike_data$rideable_type == "electric_bike" & combined_bike_data$member_casual == "member")
sum(combined_bike_data$rideable_type == "electric_bike" & combined_bike_data$member_casual == "casual")
sum(combined_bike_data$rideable_type == "classic_bike" & combined_bike_data$member_casual == "member")
sum(combined_bike_data$rideable_type == "classic_bike" & combined_bike_data$member_casual == "casual")
sum(combined_bike_data$rideable_type == "docked_bike" & combined_bike_data$member_casual == "member")
sum(combined_bike_data$rideable_type == "docked_bike" & combined_bike_data$member_casual == "casual")
```


# Visualize

### Comparing Members Vs Casual Riders

#### Day of Week

Using the `combined_bike_data` data frame we can see that members increase using their bikes on Thursdays and casual riders are using bikes more on Saturdays, this shows that there is a difference in the riding behavior between members and casual riders

```{r}
member_casual_color <- c (
"casual" = "brown",
"member" = "black"
)

combined_bike_data %>%
  group_by(member_casual, day_of_week) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = day_of_week, y = count, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(count, 2)), position = position_dodge(width = 1.1), vjust = -0.5, size = 3) +
  scale_fill_manual(values = member_casual_color, c("Member_Casual"))+
  labs(x = "Day Of Week", y = "Member-Casual Count", title = "Member vs Casual Riders By Week Days")+
  theme_minimal()
```

#### Month

Using the same inputs from the last visualization but changing the `group_by` to group it by `month`.
We can see from the visualization that for `members` August has the highest member riders for the past 12 month and `Casual` riders are riding bikes the most in July, overall June, July, August, and September are the highest 4 month of the past 12 that have the highest riders members and casual riders. 
The lowest month with riders are December, January, February, and March, they have the lowest riders for the past 12 month, this means that riders either members or casual riders usually ride the most in summer and ride fewer in the winter.

```{r}
combined_bike_data %>%
  group_by(month, member_casual) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month, y = count, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(count, 2)), position = position_dodge(width = 0.7 ), vjust =    -0.5 ,size = 3, color = "darkgrey") +
  scale_fill_manual(values = member_casual_color,c("Member_Casual"))+
  labs(x = "Month", y = "Member-Casual Count", title = "Member Vs Casual Rides By Month")+
  theme_minimal()
```

### Comparing Classic Bike vs Electric Bike

This visualization shows comparison between classic bikes and electric bikes based on usage by casual riders and members.
As you can see that the classic bikes are being used more by both members and casual riders but the overall usage of bike is more by members.

```{r}
combined_bike_data %>%
  filter(rideable_type %in% c("classic_bike", "electric_bike")) %>%
  group_by(rideable_type, member_casual) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = rideable_type, y = count, fill = member_casual))+
  geom_col(position = "dodge") +
  geom_text(aes(label = round(count, 2)), position = position_dodge(width = 0.7 ), vjust = -0.5 ,size = 3) +
  scale_fill_manual(values = member_casual_color, c("Member_Casual")) + 
  labs(x = "Rideable Type", y = "Member-Casual Count", title = "Classic Bike Vs Electric Bike") +
  theme_minimal()
```

### Ride Length

#### AVG Ride Length

comparing members vs casual riders based ob=n their ride length to see what are their riding habits. 
This visualization shows that the casual riders tend to spend more time riding as their avg rides exceed 22.92 minutes, members spend much less time as their avg rides length is shorter than 12.23 minutes.

```{r}
combined_bike_data %>%
  group_by(member_casual) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = member_casual, y = avg_ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 2)), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = member_casual_color, c("Member_Casual"))+
  labs(x = "Member_Casual", y = "Ride Length", title = "Member vs Casual Riders By Ride Length")+
  theme_minimal()
```

#### AVG Ride Length by Weekday

Comparing the avg `ride_length` by weekdays and `rideable_type`.
for casual riders the longest they ride are on Sundays with more than an average of 26.8 minutes and members as well are on Sundays but with an average of almost 13.71 minutes.
the shortest they ride are on Wednesdays for casual riders with and average of 19.62 minutes and Mondays and Thursdays for members with an average of 11.68 minutes.

```{r}
combined_bike_data %>%
  group_by(day_of_week, member_casual) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = day_of_week, y = avg_ride_length, fill = member_casual)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 2)), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = member_casual_color,c("Member_Casual"))+
  labs(x = "Day Of Week", y = "Average Ride Length", title = "Average Ride Length by Weekday and Member_Casual")+
  theme_minimal()
```

#### AVG Ride Length and Rideable Type

Comparing the avg ride length based on the rideable type.
The visualization shows that the classic bike has an average length of 17.17 minutes and the electric bike has and average of 12.14 minutes, which means that the classic bike is used more for long trips and the electric bikes are for shorter rides.

```{r}
classic_electric <- c (
  "classic_bike" = "brown",
  "electric_bike" = "black"
)
combined_bike_data %>%
  filter(rideable_type %in% c("classic_bike", "electric_bike")) %>%
  group_by(rideable_type) %>%
  summarise(avg_ride_length = mean(ride_length, na.rm = TRUE)) %>%
  ggplot(aes(x = rideable_type, y = avg_ride_length, fill = rideable_type)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(avg_ride_length, 2)), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = classic_electric)+
  labs(x = "Rideable Type", y = "Average Ride Length", title = "Average Ride Length by Rideable Type")+
  theme_minimal()
```
